# Loss distributions

```{r 02-loss-distributions-01, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Learning Objectives {#id-02-learning-objectives}

1. Describe the properties of the statistical distributions which are suitable for modelling individual and aggregate losses.
2. Explain the concepts of excesses, deductibles and retention limits.
3. Describe the operation of proportional and excess of loss reinsurance.
4. Derive the distribution and corresponding moments of the claim amounts paid by the insurer and the reinsurer in the presence of excesses (deductibles) and reinsurance.
5. Estimate the parameters of a failure time or loss distribution when the data is complete, or when it is incomplete, using maximum likelihood and the method of moments.
6. Fit a statistical distribution to a dataset and calculate appropriate goodness of fit measures.

## Theory {#id-02-theory}

**TO ADD THEORY ABOUT LOSS DISTRIBUTIONS HERE**

### Probability distributions for modelling insurance losses

### Mechanisms for limiting insurance losses

### Proportional and Excess of Loss reinsurance

### Estimating parameters of loss distributions with complete data

### Estimating parameters of loss distributions with incomplete data

## `R` Practice {#id-02-practice}

We are investigating the reinsurance arrangement of 1,000 insurance claims named `X` with the following characteristics:

* $X \sim Exp(0.01)$
* Unlimited excess of loss reinsurance, with a retention level of $M = 400$

```{r 02-loss-distributions-02, message=FALSE}
library(dplyr) # Data manipulation

set.seed(1027) # Fix result
X <- rexp(
  n = 1000,
  rate = 0.01
)

M <- 400

summary(X)
```
We want to determine the proportion of claims that are fully covered by the insurer:

```{r 02-loss-distributions-03}

Proportion <- sum(X <= M) / length(X)

```

The proportion of claims that are fully covered by the insurer is `r paste0(round(Proportion * 100,3),"%")`.

Next, for each claim, we want to calculate the net (_of reinsurance_) amount paid by the insurer. We will record this in a vector called `Y`:

```{r 02-loss-distributions-04}
Y <- ifelse(X > M, M, X)

summary(Y)
```

Likewise, for each claim, we want to calculate the amount paid by the **re**insurer. We will record this in a vector called `Z`:

```{r 02-loss-distributions-05}
Z <- ifelse(X > M, X - M, 0)

summary(Z)
```
Now let us assume that the underlying gross claims distribution follows an exponential distribution of some unknown rate $\lambda$. We will estimate $\lambda$ using only the retained claim amounts which we have recorded in vector `Y`.

First let's calculate the log-likelihood as a function of the parameter $\lambda$ and claims data `Y`:

```{r 02-loss-distributions-06}
#TO DO
```

We now determine the value of $\lambda$ at which the log-likelihood function reaches its maximum:

```{r 02-loss-distributions-07}
#TO DO
```

Finally let's plot the results:

```{r 02-loss-distributions-08}
library(ggplot2)
#TO DO
```
